---
title: "Exploring Equity in Twin Cities: A Spatial Analysis of Ridership and Income"
---

```{r setup}
#| label: setup
#| include: false
library(tidyverse)
library(dplyr)
library(sf)
library(tidycensus)
library(patchwork)
library(scales)
```

```{r load-data}
#| label: load-data
#| include: false
#| cache: true

# Transit shapefile
df <- st_read("data/raw/shp_trans_transit_count_headway_sum/", quiet = TRUE)

# ACS median household income for MN tracts
tracts <- get_acs(
  geography = "tract",
  variables = c(
    median_income = "B19013_001",   
    total_pop     = "B02001_001", 
    white         = "B02001_002", 
    black         = "B02001_003",   
    native        = "B02001_004",  
    asian         = "B02001_005",  
    hispanic      = "B03003_003"   
  ),
  state = "MN",
  year = 2022,
  geometry = TRUE
)
```

```{r process-data}
#| label: process-data
#| include: false

# Reproject transit lines to match ACS CRS
lines <- st_transform(df, st_crs(tracts))

# Spatial join: assign each transit segment to the tract it intersects
lines_tracts <- st_join(lines, tracts)

# Summarise trip count and income per tract
tract_summary <- lines_tracts %>%
  group_by(GEOID) %>%
  summarise(
    total_trips = sum(TripCount, na.rm = TRUE),
    median_income = first(estimate)
  )

# Join summaries back to tract geometries and filter to 7-county metro
tracts_map <- tracts %>%
  left_join(st_drop_geometry(tract_summary), by = "GEOID")

metro_tracts <- tracts_map %>%
  filter(grepl("Hennepin|Ramsey|Dakota|Anoka|Washington|Scott|Carver", NAME)) %>%
  mutate(county = str_extract(NAME, "Hennepin|Ramsey|Dakota|Anoka|Washington|Scott|Carver"))
```

## Motivation

Public transit is one of the most direct ways cities can expand economic opportunity. It determines whether someone can reach work, school, a doctor, or a grocery store without a car. In a car-dependent metro like the Twin Cities, who has access to frequent, reliable transit matters.

Our motivation for this project came from wanting to test a assumption: does Metro Transit
actually serve the people who need it most? The Twin Cities is often considered one of the most economically segregated metros in the country, and transit equity is an important policy debate. We had spatial data to investigate the issue directly, which led us to the central question: does the transit network follow need, or does it follow
something else?

This project matters because transit access is a mechanism for economic mobility. Understanding where service is strong, where it falls short, and who bears the burden of those gaps is essential for advocates, planners, and policymakers working toward a more equitable region.

---

## Research Question

**Does Twin Cities transit adequately serve marginalized communities?**

This overarching question breaks into three testable sub-questions, each corresponding
to a thread of our analysis:

1. Is there a relationship between neighborhood income and transit service volume?
2. Do communities of color have greater or lesser access to transit routes?
3. How does service quality (measured by headways and time-of-day coverage) vary
   across the network, and what does that mean for riders who depend on off-peak
   service?

The key idea we want to communicate: **having a bus route nearby is not the same as
having good transit.** Frequency, hours of operation, and wait times matter as much
as geographic coverage.

---

## Background

Before diving into the analysis, a few concepts and terms need clarification.

**Census tracts** are small geographic units maintained by the U.S. Census Bureau,
typically containing between 1,200 and 8,000 people. They serve as the primary unit
of analysis throughout this report, allowing us to compare transit service levels
across neighborhoods.

**Headway** refers to the number of minutes between consecutive trips on a given
route segment. A headway of 10 minutes means a bus arrives every 10 minutes; a
headway of 60 means riders may wait up to an hour. Headway is arguably the most
important measure of transit quality — it determines whether transit is a genuinely
viable option or a last resort.

**Service periods** in the dataset are broken into seven windows:

| Period     | Abbreviation | Time Range          |
|------------|-------------|---------------------|
| Early AM   | EAM         | 4:30 – 6:00 AM      |
| AM Peak    | AMP         | 6:00 – 9:00 AM      |
| Midday     | MID         | 9:00 AM – 3:00 PM   |
| PM Peak    | PMP         | 3:00 – 6:30 PM      |
| Evening    | EVE         | 6:30 – 9:00 PM      |
| Night      | NIGHT       | 9:00 PM – 1:30 AM   |
| Owl        | OWL         | 1:30 – 4:30 AM      |

**The 7-county metro** refers to Hennepin, Ramsey, Dakota, Anoka, Washington, Scott,
and Carver counties — the standard definition of the Twin Cities metropolitan area
used by the Metropolitan Council.

**ACS (American Community Survey)** is an ongoing U.S. Census Bureau survey that
provides annual demographic and economic estimates — including median household
income and racial composition — at the census tract level.

**TripCount** is a field in the transit dataset representing the total number of
scheduled transit trips serving a given road segment across all service periods.
Higher values indicate more frequent service on that segment.

---

## Data

### Data Collection

We used two datasets in this project.

The **Metro Transit Count and Headway Summary** shapefile comes from the Minnesota
Geospatial Commons, produced by the Metropolitan Council's GIS batch process
(`SVC-GISBATCH`). The dataset was last updated March 21, 2026. It contains 75,634
road segment features, each representing a stretch of road served by Metro Transit,
with trip counts and headway values recorded for each of the seven service periods
across three day types: weekday (`WK`), Saturday (`SAT`), and Sunday (`SUN`).
Geometry is stored as multiline strings in the NAD83/UTM Zone 15N projected
coordinate reference system.

The **American Community Survey (2022)** data was pulled via the `tidycensus` R
package for all Minnesota census tracts, using variable `B19013_001` (median
household income), with tract polygon geometries included. This data is collected
continuously by the Census Bureau and represents 5-year estimates centered on 2022.

### Data Acquisition

The transit shapefile was downloaded directly from the Minnesota Geospatial Commons:

> <https://gisdata.mn.gov/dataset/us-mn-state-metc-trans-transit-count-headway-sum>

Census tract data was accessed via the Census Bureau API using `tidycensus::get_acs()`
with `geometry = TRUE`, requiring a Census API key.

### Data Understanding

The transit shapefile contains **75,634 features** and **23 attribute fields**, including
time-of-day trip counts (`EAM_count`, `AMP_count`, etc.), headway values
(`EAM_hdwy`, `AMP_hdwy`, etc.), a `service_id` field (day type), and road segment
identifiers. The `TripCount` field aggregates trips across all periods for a given
segment and day type.

Before analysis, several preparation steps were required:

- **CRS reprojection:** The transit lines were in NAD83/UTM Zone 15N (EPSG:26915),
  while the ACS tract geometries used geographic NAD83 (EPSG:4269). We reprojected
  the transit lines to match the ACS CRS using `st_transform()` before performing
  any spatial operations.
- **Spatial join:** We used `st_join()` to match each transit line segment to the
  census tract polygon it intersects, then summarized total trips and median income
  per tract via `group_by(GEOID) |> summarise()`.
- **Metro filter:** We filtered to the 7-county metro area using a `grepl()` match
  on county names in the tract `NAME` field, yielding **784 census tracts**.
- **Missing data:** 177 metro tracts had missing median income estimates from the
  ACS. These tracts were retained in spatial visualizations but excluded from
  income-based scatter plots and regressions. Missing values tend to occur in
  group-quarters geographies, parks, and industrial zones.
- **Skewness:** `TripCount` values were right-skewed (min: 1, median: 31, mean: 49,
  max: 836), so `log1p()` transformation was applied in choropleth maps and
  scatterplots to reduce the influence of extreme values.

A key quality check confirmed that **zero transit line segments failed to match a
census tract** — every segment fell within a valid geography. The histogram below
illustrates the skewness that motivated log transformation.

```{r trip-count-distribution}
ggplot(lines_tracts, aes(x = TripCount)) +
  geom_histogram(bins = 60, fill = "#2c7bb6", color = "white", linewidth = 0.2) +
  scale_x_continuous(labels = comma) +
  labs(
    title = "Distribution of Transit Trip Counts per Segment",
    x = "Trip Count", y = "Number of Segments"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```
---

## Data Insights

### Transit service concentrates in lower-income tracts

Our analysis began with the map below, which overlays Metro Transit routes on a choropleth of median household income across the 7-county metro. Transit lines cluster densely in the urban cores of Minneapolis and Saint Paul, precisely where incomes are lowest. Moving outward into wealthier suburbs, the map darkens in income and empties of transit lines almost entirely.

This inverse relationship raises an important distinction. The network is reaching lower-income communities geographically, but presence alone does not tell us whether service is frequent enough, or available at the right hours, to function as reliable transportation. 

```{r}
ggplot() +
  geom_sf(data = metro_tracts, aes(fill = estimate), color = "white", size = 0.1) +
  scale_fill_gradient(
    name = "Median Income ($)",
    low = "#d4f1d4",
    high = "#1a5c1a",
    labels = scales::dollar_format()
  ) +
  geom_sf(data = lines, size = 0.2, alpha = 0.6) +
  coord_sf(
    xlim = c(-94.0, -92.7),
    ylim = c(44.6, 45.3)
  ) +
  labs(
    title = "Transit Routes vs. Median Household Income"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", size = 15, hjust = 0.5)
  )
```




The pattern can be further observed in the side-by-side choropleth maps comparing median household income to log-scaled transit trip density which reveal a clear spatial inverse: the
urban cores of Hennepin and Ramsey counties — which have the lowest
median incomes and highest concentrations of people of color — have by
far the most transit service. The wealthiest outer-ring counties are
near-blank on the transit map.

```{r choropleth-maps}
common_theme <- theme_void(base_size = 10) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 13),
    legend.position = "bottom",
    legend.key.width = unit(1.5, "cm")
  )

p1 <- ggplot(metro_tracts) +
  geom_sf(aes(fill = median_income), color = "white", linewidth = 0.1) +
  scale_fill_viridis_c(labels = dollar, name = "Median Income", na.value = "grey85") +
  labs(title = "Median Household Income") +
  common_theme

p2 <- ggplot(metro_tracts) +
  geom_sf(aes(fill = log1p(total_trips)), color = "white", linewidth = 0.1) +
  scale_fill_viridis_c(option = "magma", name = "Transit Trips (log)", na.value = "grey85") +
  labs(title = "Transit Trip Density") +
  common_theme

p1 + p2 +
  plot_annotation(
    title = "Income vs. Transit Service Across Twin Cities Census Tracts",
    theme = theme(plot.title = element_text(face = "bold", size = 15, hjust = 0.5))
  )
```

The county-level bar chart reinforces the pattern. Carver County — the
wealthiest in the metro — averages just 334 trips per tract, compared to
over 9,000 in Hennepin. Bar color encodes mean income, making the
income–service inverse immediately visible.

```{r county-summary}
county_summary <- metro_tracts %>%
  st_drop_geometry() %>%
  group_by(county) %>%
  summarise(
    mean_trips  = mean(total_trips, na.rm = TRUE),
    mean_income = mean(median_income, na.rm = TRUE),
    n_tracts    = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_trips))

ggplot(county_summary, aes(x = fct_reorder(county, mean_trips), y = mean_trips)) +
  geom_col(aes(fill = mean_income), width = 0.7) +
  geom_text(aes(label = comma(round(mean_trips))), hjust = -0.1, size = 3.5) +
  scale_fill_viridis_c(labels = dollar, name = "Mean Median Income") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  coord_flip() +
  labs(
    title = "Mean Transit Trips per Tract by County",
    subtitle = "Bar color reflects average median household income",
    x = NULL, y = "Mean Trip Count"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom",
    legend.key.width = unit(1.5, "cm")
  )
```

This pattern is consistent with how most American cities are structured:
higher-density, lower-income urban neighborhoods are more
transit-oriented by design, and investment has historically followed
those corridors.

### Transit service also concentrates in high-POC urban tracts

A parallel pattern emerges when race is substituted for income. The map below overlays Metro Transit routes on the share of people of color by census tract. The densest transit corridors run directly through the tracts with the highest concentrations of people of color, the dark purple core spanning Minneapolis and Saint Paul. Suburban tracts, overwhelmingly lighter in shade, receive comparatively little service.

What the map cannot tell us, however, is whether the network was designed with these communities in mind, or whether the alignment is largely a byproduct of density and historical land use. That distinction matters for policy: correlation is not the same as intentional investment.

```{r}
metro_tracts_wide <- metro_tracts %>%
  select(GEOID, variable, estimate, geometry) %>%
  pivot_wider(names_from = variable, values_from = estimate) %>%
  mutate(percent_poc = ((total_pop - white) / total_pop) * 100)

ggplot() + geom_sf(data = metro_tracts_wide, aes(fill = percent_poc), color = "white", size = 0.1) + scale_fill_gradient(
  name = "% People of Color",
    low = "#f7f4ff",
    high = "#4a1486",
  labels = scales::percent_format(scale = 1)) +
  geom_sf(data = lines, size = 0.3, alpha = 0.6) +
  coord_sf(
    xlim = c(-94.0, -92.7),
    ylim = c(44.6, 45.3)
  ) +
  labs(
    title = "Transit Routes vs. Median Household Income"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", size = 15, hjust = 0.5)
  )
```

```{r}
ggplot(tract_summary, aes(x = percent_poc, y = log1p(total_trips))) +
  geom_point(alpha = 0.35, color = "#6a0dad", size = 1.5) +
  geom_smooth(method = "lm", color = "#d73027", linetype = "dashed",
              se = TRUE, linewidth = 0.9) +
  scale_x_continuous(labels = percent_format()) +
  labs(
    title = "Transit Service vs. Percent People of Color",
    subtitle = "Each point = one census tract · Red dashed = linear fit",
    x = "% People of Color",
    y = "Transit Trips (log scale)"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```
The scatter plot puts numbers to what the map suggested. A positive relationship exists between the share of people of color in a tract and transit trip volume, but the fit is weak and the scatter wide. Many high-POC tracts have modest service, and some moderate-POC tracts near major corridors have substantial coverage.

The takeaway is narrow but honest: the network is not avoiding communities of color, but the alignment is not strong enough to claim it was designed around them either.


### "Served" does not mean "well-served"

The spatial picture is only half the story. The temporal analysis reveals that
**the network is built around peak-hour commuters**, and service quality degrades
sharply outside those windows. The chart below shows how headways lengthen
dramatically as the day moves away from peak commute times.

```{r headway-by-period}

headway_long <- df %>%
  st_drop_geometry() %>%
  filter(service_id == "WK") %>%
  select(ends_with("_hdwy")) %>%
  pivot_longer(everything(), names_to = "period_raw", values_to = "headway") %>%
  filter(!is.na(headway), headway > 0) %>%
  mutate(
    period = str_remove(period_raw, "_hdwy") %>% str_to_upper(),
    period = factor(period, levels = c("EAM", "AMP", "MID", "PMP", "EVE", "NIGHT", "OWL"))
  )

headway_summary <- headway_long %>%
  group_by(period) %>%
  summarise(
    median_hdwy = median(headway, na.rm = TRUE),
    mean_hdwy   = mean(headway, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(headway_summary, aes(x = period, y = median_hdwy, fill = median_hdwy)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  geom_text(aes(label = paste0(round(median_hdwy), " min")),
            vjust = -0.4, size = 3.8, fontface = "bold") +
  scale_fill_gradient(low = "#2ca25f", high = "#d73027") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Median Headway by Service Period (Weekday)",
    subtitle = "Higher values = longer waits between buses",
    x = "Service Period", y = "Median Headway (minutes)"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

Key takeaways:

- Median headways during AM and PM Peak are **~30 minutes** — already lengthy by
  transit standards.
- Median night headways (9:00 PM – 1:30 AM) rise to **45 minutes**, with a mean
  exceeding 90 minutes.
- Owl service has a median headway of **180 minutes** — one bus every three hours.
- Only **56.4%** of weekday segments have any night service; just **9.3%** have Owl
  service, concentrated along a handful of core corridors.

Workers who depend on transit for evening or overnight shifts — disproportionately
lower-wage workers in service, healthcare, and hospitality — face a dramatically
different system than peak-hour commuters.

### Weekend service is smoother but thinner

Weekday trip volumes (~1.54 million total) substantially exceed Saturday (~1.16M) and
Sunday (~1.02M). More importantly, **the shape of service differs**: weekday trips
show a strong dual-peak structure reflecting 9-to-5 commute patterns, while weekend
service is more evenly distributed across the day. This matters for equity —
lower-income riders are more likely to work non-standard hours and depend on weekend
transit, which covers fewer routes with wider headways.

```{r trips-by-daytype}

trips_by_day <- df %>%
  st_drop_geometry() %>%
  select(service_id, starts_with("EAM_count"), starts_with("AMP_count"),
         starts_with("MID_count"), starts_with("PMP_count"),
         starts_with("EVE_count"), starts_with("NIGHT_count"),
         starts_with("OWL_count")) %>%
  pivot_longer(-service_id, names_to = "period_raw", values_to = "count") %>%
  filter(!is.na(count)) %>%
  mutate(
    period = str_remove(period_raw, "_count") %>% str_to_upper(),
    period = factor(period, levels = c("EAM", "AMP", "MID", "PMP", "EVE", "NIGHT", "OWL")),
    day_type = case_when(
      service_id == "WK"  ~ "Weekday",
      service_id == "SAT" ~ "Saturday",
      service_id == "SUN" ~ "Sunday"
    )
  ) %>%
  filter(!is.na(day_type)) %>%
  group_by(day_type, period) %>%
  summarise(total = sum(count, na.rm = TRUE), .groups = "drop")

ggplot(trips_by_day, aes(x = period, y = total / 1e6, fill = day_type)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(
    values = c("Weekday" = "#2c7bb6", "Saturday" = "#fdae61", "Sunday" = "#d7191c"),
    name = "Day Type"
  ) +
  scale_y_continuous(labels = function(x) paste0(x, "M")) +
  labs(
    title = "Total Transit Trips by Period and Day Type",
    x = "Service Period", y = "Total Trips (millions)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

### The income-transit relationship is real but noisy

Both a linear regression and a LOESS smoother show a weak negative relationship
between median household income and total trip count: lower-income tracts tend to
have more service. However, the relationship has wide scatter. Some lower-income
tracts have very little service; some moderate-income tracts near high-frequency
corridors have substantial coverage.

```{r income-trips-scatter}

metro_tracts %>%
  st_drop_geometry() %>%
  filter(!is.na(median_income), !is.na(total_trips)) %>%
  ggplot(aes(x = median_income, y = log1p(total_trips))) +
  geom_point(alpha = 0.35, color = "#4d4d4d", size = 1.5) +
  geom_smooth(method = "loess", color = "#2c7bb6", fill = "#abd9e9", linewidth = 1.1) +
  geom_smooth(method = "lm", color = "#d73027", linetype = "dashed", se = FALSE,
              linewidth = 0.8) +
  labs(
    title = "Median Income vs. Transit Trips per Census Tract",
    subtitle = "Blue = LOESS smoother · Red dashed = linear fit · Each point = one census tract",
    x = "Median Household Income", y = "Transit Trips (log scale)"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

This suggests that **density, route history, and corridor geography explain service
distribution at least as much as income alone.** The network was not explicitly
designed with income equity as a primary goal, and the spatial alignment with
lower-income tracts may be largely incidental to land-use patterns.

### A surprising result worth investigating

The scatter plot of total trips vs. percent people of color shows only a weak
positive relationship — weaker than the visual map overlay would suggest. This
disconnect is likely a methodological artifact: our spatial join assigns the full
trip count of any line segment to the single tract it intersects, which can inflate
counts in tracts that are physically large or positioned at route intersections. This
is one of the more important limitations to address in future work.

---

## Conclusions

The data tells a story that is **partially encouraging and partially concerning**.

At the broadest level, transit service is concentrated where lower-income communities
and communities of color live — the network is spatially aligned with need in
relative terms. The urban cores of Minneapolis and Saint Paul are well-covered, and
the county-level gradient in service volume tracks (inversely) with income.

But spatial coverage is only half of transit equity. When we examine *when* service
runs and *how often*, a different picture emerges. The system is structured for
peak-hour, 9-to-5 commuters. Riders who need transit most — lower-wage workers,
people without cars, those working evenings and overnight — face headways that make
transit impractical for daily life. A 90-minute mean night headway is not a system
that functions as reliable transportation; it is a system that merely tolerates
being used outside of convenient hours.

The answer to our guiding question — *does Twin Cities transit adequately serve
marginalized communities?* — is: **geographically, yes, to a degree; temporally,
much less so.** The gap between where routes exist and when those routes actually
run is where equity is lost.

---

## Limitations and Future Work

Several limitations of this analysis are worth naming directly.

**Spatial join methodology.** Our `st_join()` assigns the full trip count of each
line segment to whichever tract polygon it intersects. Segments that cross tract
boundaries are counted in only one tract, which can artificially inflate or deflate
per-tract totals — particularly for large or irregularly shaped tracts near
corridors. A more rigorous approach would use `st_intersection()` to clip segments
to tract boundaries and weight trip counts proportionally by the share of segment
length within each tract.

**Missing income data.** 177 metro tracts (~22%) have missing ACS median income
estimates and were excluded from regression analyses. These missing values are not
random — they tend to occur in group-quarters geographies and institutional areas —
which may introduce systematic bias.

**Cross-sectional snapshot.** The dataset represents a single point in time (March
2026). We cannot draw conclusions about trends, or whether recent service changes
have disproportionately affected certain communities.

**Service supply vs. ridership demand.** Our analysis measures service supply
(scheduled trips per segment) rather than actual ridership. High supply in a tract
does not necessarily mean high ridership, and vice versa. Incorporating actual
boarding data from Metro Transit's ridership records would provide a more complete
picture.

**Race/ethnicity analysis not incorporated with income data** Racial demographics were analyzed but not incorporated with income data. We would have to  conduct multivariate analysis controlling for density, income, and racial composition simultaneously.

**Future directions** could include: accessibility metrics measuring how many jobs
or essential services can be reached within 30 minutes by transit from each tract;
a longitudinal analysis of service changes; disaggregation of headway equity by
income quintile; and comparison with peer cities to contextualize the Twin Cities'
equity performance.
